### 2020-02-14 response/response-demography-2.tex

\documentclass[letterpaper,12pt]{article}

% Xetex preamble
\usepackage{fontspec}
\setromanfont[Ligatures=TeX]{TeX Gyre Pagella}
\usepackage{unicode-math}
\setmathfont{TeX Gyre Pagella Math}

\usepackage[longnamesfirst]{natbib}
%\usepackage[tabhead,nolists,tablesfirst]{endfloat}
\usepackage[flushleft]{threeparttable}
\usepackage{booktabs}
\usepackage{rotating}  
\usepackage{dcolumn} 
\usepackage{setspace}
%\usepackage{flafter} 
%\usepackage{longtable}
%\usepackage[pdftex]{graphicx}
\usepackage[xetex,colorlinks=true,linkcolor=black,citecolor=black]{hyperref} 
\usepackage[margin=1.0in]{geometry} 
\usepackage{multirow}

@@

\begin{document}

\begin{center} \textbf{\large Birth Spacing in the Presence of Son
Preference and 
Sex-Selective Abortions: India's Experience over Four Decades}
\end{center}

\begin{center} Response to Editor and Reviewer Comments \end{center}

\noindent Please find attached the revised version of my paper,
``Birth Spacing in the Presence of Son Preference and Sex-Selective
Abortions:
India's Experience over Four Decades.''
I want to thank the three reviewers and the editor for the constructive comments and 
suggestions, which I believe substantially improved the paper.

I have substantially rewritten the paper, so before responding to the
individual comments, I provide an overview of the revised paper and the
most significant changes.

To clarify the contribution of the paper, I now focus on three
questions. First, how has birth spacing changed over time with the
introduction of sex selection and the other significant changes in
India? Second, with substantial increases in spacing, how biased is our
standard fertility estimates? Finally, how has infant mortality changed,
and is there an effect of sex selection on infant mortality risk?

I have added four new sections: 
two analytical sections, one on fertility and one on infant mortality, 
and two background sections, one on female education and labor force 
participation and one on the conceptual framework.
I have removed the section on regional differences.

Section-by-section overview of significant changes:
\begin{enumerate}

\item I have completely rewritten the ``Introduction'' section to provide background on 
India, discuss why examining birth spacing is relevant, and highlight main results.

\item The first new section covers how female education and labor force
participation changed over time in India. Both female education and
labor force participation are factors in spacing decisions. That labor
force participation has declined even with new groups received more
schooling helps address the question of whether the expanded access to
education might change behavior.

Because labor force participation differs by education level, I now divide the original 
eight or more years of education group into two groups: eight to eleven and twelve or
more years of education. The eight to eleven group has the lowest labor
force participation of any group, and twelve or more has one of the
highest.

\item The second new section provides a conceptual framework.
I first work through the likely general changes based on the increasing female education 
and the declining labor force participation and second through the potential effects of 
access to sex selection.
I end with a summary of the predictions.

\item I have further shortened the ``Estimation Strategy'' section and
edited it to focus on the motivation for the choice of model.

\item The Data section is also shorter, and the discussion of potential
effects of variables is now in the conceptual framework section.

\item With the new education grouping, I have completely rewritten the
section on how birth spacing has changed (the original Results section).
Instead of going through each education level in detail, I now focus
more broadly on how average birth spacing and parity progression have
changed over time and then on the effects of sex selection, drawing on
examples to illustrate the changes. The graphs are now easier to read.

\item The original subsection ``Distribution of Birth Spacing Within
Spells'' is now a new separate section ``Distribution of Birth Spacing''
and has been completely rewritten. Instead of showing the conditional
survival graphs, the section now includes figures with the 25th and 75th
percentiles for birth intervals. These figures allow me to focus on the
changes in both short and long intervals. Both to tie in with the
subsequent mortality discussion.

\item The new section, ``What has Happened to Fertility?'' looks at the
degree of tempo effect that arises from the longer spacing by comparing
a TFR-like measure with the predicted cohort fertility based on the
hazard model results.

\item The new section ``Mortality and the Changing Birth Spacing''
examines how the association between infant mortality and birth spacing
has changed over time by the education group and whether the increased
use of sex selection increases mortality risk.

\item Finally, the conclusion has been completely rewritten.
\end{enumerate}

The revised paper is just over 8,000 words per the author guidelines.

@@

\begin{enumerate}

\item First, greater explanation of the unique contributions of this
study is needed. As the reviewers note, the literatures in the areas of
son preference and birth spacing, for example, are extensive, and
clearer and more compelling motivations for the work should be provided.

[Response:]

To better address why and how the changes in birth spacing are relevant,
I have added two new sections. One discusses the bias that arises in
fertility estimates from the significant increases in birth spacing,
both from sex selection and from secular changes, and show that we
likely have substantially overestimated how early and fast fertility
fell in India. The other shows that mortality has fallen for all groups
no matter how long birth intervals are and that there is no apparent
adverse effect of high use of sex selection on infant mortality.

@@

\item Second, the paper lacks a conceptual framework to organize (and
motivate) the analysis. Following on the comments of one reviewer, one
suggestion would be to focus on the role of women’s education, with a
more thorough description of the context of changing education in India
over the last half century and clearly lay out a framework for the
mechanisms through which birth spacing is affected by educational
expansion.

[Response:]

I address this in two parts.

First, I included a new descriptive section on how female education and
labor force participation has changed over time in India. I include the
discussion of labor force participation because theories and prior
results reveal it to be a crucial component in the link between female
education and birth spacing. I argue that ``Sanskritization'' is behind
the continued low labor force participation despite the expanded
schooling and the associated changes in the composition of the education
groups.
The implication is that ``Sanskritization'' likely carries over to other decisions, such 
as sex selection use.

Second, I added a conceptual framework that focuses on how birth spacing
is likely to respond in a situation where labor force participation is
low and declining, and son preference is strong. A more general theory
overview would be interesting, but that is not feasible given the word
constraint. The conceptual framework section also discusses predictions
for how access to sex selection is likely to impact birth spacing and
summarizes the expected effect of the changes in India on birth spacing.

Writing these sections also showed that my original classification of
education was too broad because labor force participation differs
markedly within the original eight-plus education group. I have,
therefore, redone all analyses in the paper with four education groups.

@@

\item This relates to a third issue, which is space limitations and
decisions regarding what to include in the manuscript. In the last
decision letter you were asked to “reorganized to focus on the most
important issues.” Although the author has done a good job of focusing
the paper in the revision, there are still questions about lack of
detail in some places and perhaps too much in others. One suggestion is
to drop the analysis on regional differences and focus the paper more
centrally on the topic of maternal education, for example (as noted
above).

[Response:]

I agree that the regional analysis added little to the paper, and 
have deleted this section. What I consider the two most
important issues related to birth spacing are fertility and mortality.
I, therefore, now focus on how both relate to spacing, in addition to
examining how birth spacing changed.

There are other relevant questions, such as the effect on the mother
that I have ignored because of space considerations.

As part of the rewrite, I have removed any remaining discussion that
might make the paper seem more like a methodological paper. For example,
I have shortened the ``Data'' and ``Estimation Strategy'' sections and
rewritten the description of the method in the ``Introduction'' section.

@@

\begin{enumerate}

\item The model is well explained, but the paper could be more explicit
on why a model is needed. Some of the conclusions drawn from average
probabilities, sex ratios, etc. (particularly those not conditioning on
specific values of the covariates) might also be directly visible in the
raw data. If so, the data description part could perhaps be extended.

[Response:]

Space limitations mean that I cannot add a more detailed examination of
the raw data to the data description. I have instead tried to address
this in three ways. First, I have added a discussion in the ``Estimation
Strategy'' section of why hazard models are the preferred way of dealing
with censoring. Second, I highlight that censoring of birth spells
increases with parity and time in the ``Data'' section. Third, I added a
comparison to the birth spacing numbers from the NFHS reports, which do
not attempt to deal with censoring and, therefore, have shown virtually
no change over time.

@@

\item Details on the estimation results are not presented. I assume
there are too many of them to be included in the paper. Still, the paper
extensively discusses model specification issues such as the
proportionality assumption and the way that is handled by splitting the
sample. I would like to see some discussion of model specification tests
informing the reader whether this is necessary and whether the resulting
flexible specifications indeed provide a good fit to the data.

[Response:]

I have tried to move away from pitching the paper as a methodological
contribution. As part of this, I shortened the empirical model
discussion and removed any mention in the ``Conclusion'' section. I
agree that a more in-depth examination of how a proportional model stack
up against the non-proportional model would be fascinating. However, I
believe this is best done separately, given the already considerable
length of the paper and the Appendix.

Furthermore, the non-proportionality assumption makes the model more
flexible than it would have been without, but cannot introduce any bias
compared to the proportional model. If the non-proportionality
assumption is not needed, the only thing affected is the efficiency of
the estimates. However, the reverse is not the case, with bias possible
if proportionality does not hold.

@@

I agree and have removed the section, as discussed above.

\item The concluding section raises some interesting points but these
are only indirectly related to the content of the paper. The development
of family planning possibilities with increasing availability and use
over time seems to be important for understanding the general decline in
fertility and lengthening of birth intervals. It is mentioned in the
introduction but not in the conclusions. The relation with labor market
participation may be of interest, but what is said about it was not
clear to me. Why would longer birth intervals disrupt labor force
participation (p. 32)

[Response:]

Unfortunately, the questions on labor force participation are not
detailed enough to allow a direct analysis using this data. I have
instead included a discussion of how female labor force participation
has changed over time in the two new background/conceptual framework
sections. Furthermore, I now directly discuss the prior literature and
theory on labor force participation and birth spacing in the ``Birth
Spacing: Mechanisms and Prior Findings'' section.

While one might expect that increasing income and female education would
also lengthen average birth spacing through better access and knowledge
of contraception, the empirical relationship between contraception use
and birth spacing is ambiguous 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.
Furthermore, the ability to successfully use ``low efficacy''
contraceptive methods is increasing in education, while there is no
difference by schooling when it comes to modern contraceptives
\citep{Rosenzweig1989}.
Because there is not sufficient historical information on contraception
use in the data and that the prior literature does not provide a clear
relationship, I have not included this discussion in the paper.
\end{enumerate}

\noindent Specific comments:

@@

I have changed this throughout the paper.

\item p.7 after eq. 3: Sex composition of previous children is in
principle endogenous  if there is sex selective abortion (it is
essentially a lagged dependent variable). This is not taken into
account. This is, at least in theory, a limitation of the current model
that should be acknowledged. (I do not think it is easy to remedy
without explicitly incorporating unobserved heterogeneity.)

[Response:]

That is correct and something that would be interesting to address in a future paper.
For now, I have included a comment on this in the Estimation Strategy.

@@

\item p.7: Why use intervals of three months? Did the author(s) perform
any sensitivity checks?

[Response:]

I tried shorter than three months, but because the method requires
events to happen in each period, the estimations would often fail to
converge. For the same reason, I combine some three months intervals to
ensure enough information, especially when intervals are long and the
likelihood of a next birth low.

@@

\item p.9: It might be useful to already mention that Hindus only are
analyzed in the introduction, since religion might play an important
role when choosing sex selective abortion.

[Response:]

I have included this in the rewritten ``Introduction'' section and the abstract.

@@

\item p.10: Sterilization also seems an endogenous decision in this
context, this should be acknowledged as an additional limitation. How
often does this occur? (It might even be possible to treat this as an
additional competing risk.)

[Response:]

Across the entire sample, 36 percent of women report sterilization for
either themselves or their husbands. The decision does, however, not lend
itself easily to incorporation into the model as an additional competing
risk. First, almost nobody is sterilized after their first birth (only
around 0.5 percent), although this goes up to 19 percent after the
second birth, and 27 percent after the third. Second, a majority---often
up to 75 percent---of sterilizations take place in the same month as the
last birth or within eight months of it. Most of the sterilizations are,
therefore, never included in the model samples because those begin at
nine months after the prior birth. Hence, the absolute numbers are too
small for the competing risk model to work correctly, primarily because
the likelihood of sterilization depends strongly on the sex composition
of prior children. For a given parity, the fewer boys, the lower the
probability.

The results presented here are conditional on a prior birth and no
sterilization within the nine months after the preceding birth. The main
concern is that the parity progression probabilities are too high
because they do not account for sterilization. There are two impacts of
this.

First, it biases downward the differences in the parity progression
probabilities across sex compositions. I discuss this in the
``Data'' section.

Second, the predicted cohort fertility would be too high. I, therefore,
estimate the likelihood of sterilization within the first nine months
using a Logit model (results not shown in the paper). I then predict the
probability of not getting sterilized and use that to scale down the
parity progression probability when predicting cohort fertility. In
practice, the effect of the scaling is relatively small, 0.05 to 0.25,
which is consistent with the lowest use among women who have the highest
estimated parity progression probabilities based on the hazard model.

@@

\item p.9: why put the marriage duration threshold at 22 years?
Sensitivity check?

[Response:]

I discuss this in the Appendix section ``Recall Error and the Sex
Ratio''. Since most of the surveys start showing significantly biased
sex ratio from around 22 years of marriage on, I drop all observations
where the wedding took place 22 years or more. Changes of a couple of
years more or less do not substantially change the results. However, a
cut significantly higher than 22 does eventually make the sex ratio very
uneven in the first period, even for less-educated women. There are no
effects on the more recent periods.

@@

\item Results: The many small figures are hard to read and I did not
find them an attractive way to present the results. Perhaps the
author(s) can come up with an alternative?

[Response:]

I have tried to address this by increasing the font size for the legends
and turning the graphs sideways. Originally the results were presented
in table format (now available in the Appendix). Prior referees found it
hard to get an overview of the results that way, and, on balance, I
agree with that. Hence, I have kept the graphs for the spacing
discussions. For the fertility section, I present the results as a
table.

@@

This study uses four waves of the National Family and Health Surveys to
examine the interrelationship between son preference, sex ratios at
birth, and birth spacing in India, how this varies between different
regions, by parity and the sex composition of the children in the
household, by maternal educational level, and over time.

\begin{enumerate}

\item On the one hand this feels like a clear and polished piece of
research. On the other hand, I wonder how much we can learn from this
study given that there is already an extensive body of research that
examines son preference and SRBs in India, including how this varies by
parity, sex composition of the children already in the household,
maternal educational level, and over time. It seems to me that the
question is, what is the net contribution of adding birth spacing as an
additional dimension to this whole discussion.

The authors attempt to address this point by offering four motivations
for why we should care about spacing as an additional dimension. First,
they say that “researchers have made extensive use of birth spacing as a
measure of son preference, and it is critical to understand to what
extent spacing is still a useful measure of son preference” (page 2). To
my mind, the authors do not offer a particularly compelling explanation
anywhere in the paper for why it is ‘critical to understand’ whether
spacing remains a useful measure of son preference. Understanding son
preference and how that is linked to sex-selective abortions and
differential stopping behaviour (and gender dynamics more generally), is
indeed critical – but I don’t really see that the critical importance of
understanding spacing in relation to this.

@@

[Response:]

I have reframed the paper to be about birth spacing more
generically (see also my response below) and tied it to two issues where
changes in spacing have direct relevance: fertility estimates and infant mortality.
I have removed most of the discussion of spacing as a measure of son preference,
except to note that we now have cases where son preference results in longer
spacing without boys than with because of sex selection.

% What also makes birth spacing in India interesting is that opposite of what we
% have seen elsewhere, birth spacing has increased over time with development.
% The increases is especially pronounced because of the use of sex selection.

@@

\item The second and third motivations offered by the authors are the
potential consequences of (short) birth spacing for maternal health,
child health, and long-term child development. For example, the authors
cite literature showing that short spacing can have severe consequences
for child outcomes, particularly in low-income settings. However,
understanding the potential consequences of birth interval length is
much more confusing when long birth intervals are a consequence of
(sex-selective) abortions. One reason why short birth intervals have a
negative effect on child outcomes is because of maternal nutrient
depletion and insufficient time to recover from the previous pregnancy.
Children born after long birth intervals where those intervals have been
punctuated by multiple abortions will not see the same benefits as a
child born after a long interval that is not punctuated by abortions
since the mother would not be fully recovering in that intervening
period. Likewise the costs and benefits for mothers themselves of
short/long spacing may be overwhelmed by the effects of the abortions
and interrupted pregnancies.

[Response:]

Thank you for this fascinating point on the potential adverse effects of
the increased use of abortions. I have included this as one of the two
main reasons why the changes in spacing are worth examining and have
introduced a new section on infant mortality and how it varies by
spacing and sex composition over time. There is, however, no apparent
adverse effect of being a son born after girls and with longer spacing
in situations with high use of sex selection. Hence, if there is a
negative effect of multiple abortions, it is more than countered by the
protective effect of higher maternal education. It is still possible
that we would see more of an impact among less-educated women, but this
group is, currently, not heavy users of sex selection.

To focus the paper, I no longer discuss the potential effects on
mothers, although I still find that an exciting topic for future
analysis. The same goes for the long-term development of children.

@@

\item The fourth and final motivation offered is that we don’t know what
determines spacing behaviour in low- and middle-income countries.
However, the arguments that the authors develop here are more about 
contraceptive use, declining fertility, and women entering the labor market. 
Those factors don’t seem to be addressed elsewhere in the paper at all.

[Response:]

As mentioned above, I now include a discussion of the changing spacing's effect on
fertility estimates.
Similarly, although I cannot directly examine the relationship between labor force
participation and spacing because of data limitations, I do discuss the low and 
falling labor force participation in India and the effects this may have on spacing.
I have removed the discussion on contraceptive use (see also my response above).

@@

\item In the end, this study offers a careful descriptive account of how birth
spacing has changed alongside sex ratios at birth, but little more. We
should indeed care about sex ratios at birth – and people do – but there
are many studies published on that topic, and it is not clear that this
one adds something important to that particular debate. If it does, the
authors have not articulated that clearly. I think that understanding
the consequences of birth spacing for maternal and infant health is also
extremely important. But the authors do not study that link empirically.
For these reasons, I find the contribution of the study rather marginal
and confusing.

[Response:]

As described above, I have substantially rewritten the paper and
introduced new analyses to illustrate better how and why examining birth
spacing is necessary, including linking birth spacing and infant
mortality directly. I am, however, open for suggestions for alternative
ways to focus the paper and the motivations.

### 2020-02-14 paper/sex_selection_spacing-ver3-demography.tex
Final version of revised paper for Demography

% For word counting uncomment line below and raggedright below
\usepackage[nolists]{endfloat}

@@

\title{Birth Spacing in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

\author{Claus C P\"ortner\\
    Department of Economics\\
    Albers School of Business and Economics\\
    Seattle University, P.O. Box 222000\\
    Seattle, WA 98122\\
    \href{mailto:claus@clausportner.com}{\texttt{claus@clausportner.com}}\\
    \href{http://www.clausportner.com}{\texttt{www.clausportner.com}}\\
    \& \\
    Center for Studies in Demography and Ecology \\
    University of Washington\\ \vspace{2cm}
    }

@@

% Demography abstract
\noindent 
Using four rounds of India's National Family and Health Surveys and a
competing-risk hazard model, I show that average birth intervals
increased over the last four decades for all education groups. The most
significant increases are among the women most likely to use sex
selection. Despite the rise in average intervals, the likelihood of very
short spacing did not change substantially. Hence, the increases come
predominantly from the longer birth intervals getting even longer. As a
result of the longer spacing, fertility rates significantly
overestimated how fast cohort fertility fell. Although cohort fertility
and the fertility rate have started to converge, the cohort fertility is
still substantially higher than the fertility rate. Furthermore, cohort
fertility is still at or above replacement level for all but the
best-educated urban women. Finally, infant mortality risk has declined
substantially over time for all groups, but fastest for the lower
education groups, who are now close to the level of women with the most
education. Short birth spacing is still associated with higher
mortality, although the difference is small for the best-educated women.
There is no evidence that the increasing use of sex selection is
associated with higher infant mortality risk.

@@

% For word counting remove comment on the line below
\raggedright

@@

I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
Using data from the four National Family and Health Surveys (NFHS), I apply
a competing risk hazard model to the birth histories of Hindu women, covering
the period 1972 to 2016.
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son 
but also works in cases without prenatal sex selection.
The empirical model allows me to predict fertility behavior taking into
account both the likelihood of parity progression and the likelihood of having a child of 
a specific sex.

@@

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell, as is assumed by the proportional hazard model.
The proportionality assumption is especially problematic for higher-order spells where 
there are substantial differences across groups both in the likelihood of progressing to 
the next birth and how soon couples want their next child if they are going to have one.

@@

The model is a discrete-time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The unit of analysis is a spell, the period from nine months after one birth to the next.
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$ when either birth or censoring of the spell 
occurs.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete-time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, capture 
individual, household, and community characteristics discussed below,
and $D_{j}(t)$ is the piece-wise linear baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not restrict the 
baseline hazard unnecessarily.

The explanatory variables in $\mathbf{Z}$, and the interactions between them, 
constitute the non-proportional part of the model:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2).
\end{equation}
$D_j(t)$ is the piece-wise linear baseline hazard, $\mathbf{Z}_1$ captures sex 
composition of previous children, and $Z_2$ captures the area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally.
However, to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and different 
periods.%
\footnote{
I discuss the choice of which variables to use for non-proportionality in more detail
in the ``Explanatory Variables'' section.
}

@@

I exclude visitors, as well as women in any of the following 
categories: never married; no gauna yet; married more than once; divorced; 
not living with husband; inconsistent age at marriage; or education information missing.  
The same goes for women who had at least one multiple births, reported giving birth 
before age 12, had a birth before marriage, or an interval between births of less than 
nine months. 
Finally, I restrict the sample to Hindus, who constitute about 80 percent of India’s 
population.%
\footnote{
If the use of sex selection differs across groups, assuming that the
baseline hazard is the same leads to bias. 
Furthermore, the other groups are each too small to estimate different
baseline hazards for each and so different that combining them into one
group would not make sense.
}

In addition to a large number of women surveyed and the long period covered, a
benefit of the NFHS is that enumerators pay careful attention to the spacing 
between births and probe for ``missed'' births.
Nevertheless, systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, remains a potential problem.
Recall error is heavily dependent on how long ago a woman was married, and I, therefore, 
drop women married 22 years or more.%
\footnote{
See the discussion in the Appendix.
}
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

@@

The spells all begin nine months after the previous birth, which is the earliest we should 
expect to observe a new birth.
A spell continues until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place,
sterilization of the woman or her husband,
or too few births are observed for the method to work.
For all spells, censoring is set at 96 months (eight years) after a woman can first
give birth in the spell.
With these cut-offs, less than one percent of observed births occur after
the spell cut-off.

@@

The predicted parity progression probability is the likelihood of giving birth by 96 
months after the beginning of the spell.
To find the expected average duration, I first calculate, for each woman, the probability 
of giving birth in each period $t$, which I use as weights to calculated her expected 
duration. 
I then average the individual expected durations across women using the parity progression
probabilities as weights.

@@

\begin{sidewaysfigure}
\ContinuedFloat
\rotatebox[origin=c]{90}{\small{Rural}}
\subfloat[Second Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_low_rural_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_low_rural_all} 
    \end{minipage}
} 
\subfloat[Fourth Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_low_rural_all} 
    \end{minipage}
} 
\caption{(Continued) Estimated average spell length, sex ratio, and probability of 
a next birth for women with no education}
\label{fig:bootstrap_low}
\end{sidewaysfigure}

>>>>

% For word counting uncomment line below and raggedright below
% \usepackage[nolists]{endfloat}

@@

\title{Birth Spacing in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience over Four Decades%
% \protect\thanks{%
% I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
% I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
% Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
% University of Washington, University of Aarhus, the Fourth 
% Annual Conference on Population, Reproductive Health, 
% and Economic Development, and the Economic Demography Workshop for helpful 
% suggestions and comments.
% I would also like to thank Nalina Varanasi for research assistance.
% Support for development of the method from the University of Washington Royalty 
% Research Fund and the Development Research Group of the World Bank is gratefully 
% acknowledged.
% The views and findings expressed here are those of the author and
% should not be attributed to the World Bank or any of its member countries.
% Partial support for this research came from a Eunice Kennedy Shriver National
% Institute of Child Health and Human Development research infrastructure grant,
% 5R24HD042828, to the Center for Studies in Demography and Ecology at the
% University of Washington.
% }
}

\author{}

% \author{Claus C P\"ortner\\
%     Department of Economics\\
%     Albers School of Business and Economics\\
%     Seattle University, P.O. Box 222000\\
%     Seattle, WA 98122\\
%     \href{mailto:claus@clausportner.com}{\texttt{claus@clausportner.com}}\\
%     \href{http://www.clausportner.com}{\texttt{www.clausportner.com}}\\
%     \& \\
%     Center for Studies in Demography and Ecology \\
%     University of Washington\\ \vspace{2cm}
%     }

@@

% Demography abstract
\noindent 
Using four rounds of India's National Family and Health Surveys and a
competing-risk hazard model, I show that Hindu women's average birth intervals
increased over the last four decades for all education groups. The most
significant increases are among the women most likely to use sex
selection. Despite the rise in average intervals, the likelihood of very
short spacing did not change substantially. Hence, the increases come
predominantly from the longer birth intervals getting even longer. As a
result of the longer spacing, fertility rates significantly
overestimated how fast cohort fertility fell. Although cohort fertility
and the fertility rate have started to converge, the cohort fertility is
still substantially higher than the fertility rate. Furthermore, cohort
fertility is still at or above replacement level for all but the
best-educated urban women. Finally, infant mortality risk has declined
substantially over time for all groups, but fastest for the lower
education groups, who are now close to the level of women with the most
education. Short birth spacing is still associated with higher
mortality, although the difference is small for the best-educated women.
There is no evidence that the increasing use of sex selection is
associated with higher infant mortality risk.

@@

% For word counting remove comment on the line below
% \raggedright

@@

I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
Using data from the four National Family and Health Surveys (NFHS), I apply
a competing risk hazard model to the birth histories of Hindu women, covering
the period 1972 to 2016.
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son 
but also works in cases without prenatal sex selection.
The empirical model allows me to predict fertility, taking into
account both the likelihood of parity progression and the likelihood of having a child of 
a specific sex.
Finally, I examine how infant mortality risk has changed with the changing birth spacing 
and whether sex selection affects infant mortality.

@@

Second, it is unlikely---even in the absence of prenatal sex determination---that 
characteristics, such as the sex composition of previous births, have the same effects 
throughout the entire spell, as is assumed by the proportional hazard model.
If proportionality does not hold, the results are biased.
The proportionality assumption is especially problematic for higher-order spells where 
there are substantial differences across groups both in the likelihood of progressing to 
the next birth and how soon couples want their next child if they are going to have one.

@@

The model is a discrete-time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The unit of analysis is a spell, the period from nine months after one birth to the next.
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$ when either birth or censoring of the spell 
occurs.
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete-time hazard rate $h_{ijt}$ is 
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, capture 
individual, household, and community characteristics discussed below,
and $D_{j}(t)$ is the piece-wise constant baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not restrict the 
baseline hazard unnecessarily.

The explanatory variables in $\mathbf{Z}$, and the interactions between them, 
constitute the non-proportional part of the model:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2).
\end{equation}
$D_j(t)$ is the piece-wise constant baseline hazard, $\mathbf{Z}_1$ captures sex 
composition of previous children, and $Z_2$ captures the area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally.%
\footnote{
With sex selection, the composition of prior children is, in principle, endogenous.
It is beyond the scope of this paper to develop a method for dealing with this issue.
}
However, to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and different 
periods.%
\footnote{
I discuss the choice of which variables to use for non-proportionality in more detail
in the ``Explanatory Variables'' section.
}

@@

I exclude visitors, as well as women in any of the following 
categories: never married; no gauna yet; married more than once; divorced; 
not living with husband; inconsistent age at marriage; or education information missing.  
The same goes for women who had at least one multiple births, reported giving birth 
before age 12, had a birth before marriage, or an interval between births of less than 
nine months.

Finally, I restrict the sample to Hindus, who constitute about 80 percent of India’s 
population.
If the use of sex selection differs across groups, assuming that the
baseline hazard is the same leads to bias. 
Furthermore, the other groups are each too small to estimate different
baseline hazards for each and so different that combining them into one
group would not make sense.

In addition to a large number of women surveyed and the long period covered, a
benefit of the NFHS is that enumerators pay careful attention to the spacing 
between births and probe for ``missed'' births.
Nevertheless, systematic recall error, where the likelihood of reporting a deceased 
child depends on the sex of the child, remains a potential problem.
Recall error is heavily dependent on how long ago a woman was married, and I, therefore, 
drop women married 22 years or more based on the discussion in the Appendix.
The final sample consists of 
\input{../tables/num_women.tex}
% [number] women, with [number] parity one through four births.

@@

The spells all begin nine months after the previous birth, which is the earliest we should 
expect to observe a new birth.
A spell continues until either a child is born or censoring occurs.
Censoring can happen for three reasons:
the survey takes place,
sterilization of the woman or her husband,
or too few births are observed for the method to work.%
\footnote{
Most sterilizations take place right after a woman gives birth, and,
therefore, do not show up in the samples used. 
Furthermore, sterilization depends strongly on the sex composition of prior children
with lower probabilities, the fewer boys. 
The effect is that the differences in parity progression probabilities are biased 
downwards.
}
For all spells, censoring is set at 96 months (eight years) after a woman can first
give birth in the spell.
With these cut-offs, less than one percent of observed births occur after
the spell cut-off.

@@

To find the expected average duration, I first calculate, for each woman, the probability 
of giving birth in each period $t$, which I use as weights to calculated her expected 
duration. 
I then average the individual expected durations across women using the parity progression
probabilities as weights.

The predicted parity progression probability is the likelihood of giving birth by 96 
months after the beginning of the spell.

@@

\begin{sidewaysfigure}
\centering
\ContinuedFloat
\rotatebox[origin=c]{90}{\small{Rural}}
\subfloat[Second Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell2_low_rural_all}
    \end{minipage}
} 
\subfloat[Third Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell3_low_rural_all} 
    \end{minipage}
} 
\subfloat[Fourth Spell]{
    \begin{minipage}{0.31\textwidth}
        \includegraphics[width=\textwidth]{bs_spell4_low_rural_all} 
    \end{minipage}
} 
\caption{(Continued) Estimated average spell length, sex ratio, and probability of 
a next birth for women with no education}
\label{fig:bootstrap_low}
\end{sidewaysfigure}
